---
output: github_document
bibliography: "inst/REFERENCES.bib"
csl: "inst/apa-6th.csl"
---

# ANOFA: Analyses of Frequency Data

<!-- badges: start -->
[![CRAN Status](https://www.r-pkg.org/badges/version/ANOFA)](https://cran.r-project.org/package=ANOFA)
<!-- badges: end -->

```{r, echo = FALSE, message = FALSE, results = 'hide', warning = FALSE}
cat("this will be hidden; used for general initializations.\n")
library(ANOFA)
options("ANOFA.feedback" = "none") # shut down all information
```

The library `ANOFA` provides easy-to-use tools to analyze frequency data. 
It does so using the _Analysis of Frequency datA_ (ANOFA) framework 
[the full reference @lc23b]. With this set of tools, you can examined
if classification factors are non-equal (_have an effect_) and if their
interactions (in case you have more than 1 factor) are significant. You
can also examine simple effects (a.k.a. _expected marginal_ analyses). 
Finally, you can assess differences based on orthogonal contrasts.
ANOFA also comes with tools to make a plot of the frequencies along
with 95% confidence intervals [these intervals are adjusted for pair-
wise comparisons @cgh21]; with tools to compute statistical power given
some _a priori_ expected frequencies or sample size to reach a certain
statistical power. In sum, eveything you need to analyse frequencies!

The main function is `anofa()` which provide an omnibus analysis of the 
frequencies for the factors given. For example, @lm71 explore frequencies
for attending a certain type of higher education as a function of gender:

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
w <- anofa( obsfreq ~ vocation * gender, LightMargolin1971)
summary(w)
```

A plot of the frequencies can be obtained easily with 

```{r, message=FALSE, warning=FALSE}
anofaPlot(w) 
```

Owing to the interaction, simple effects can be analyzed from the _expected marginal
frequencies_ with

```{r, message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
e <- emFrequencies(w, ~ gender | vocation )
summary(e)
```

Follow-up functions includes contrasts examinations with `contrastFrequencies()'.

Power planning can be performed on frequencies using ``anofaPower2N()`` or
``anofaN2Power()`` if you can determine theoretical frequencies.

Finally, `toRaw()`, `toCompiled()`, `toTabular()`, `toLong()` and `toWide()` 
can be used to present the frequency data in other formats.

# Installation

Note that the package is named using UPPERCASE letters whereas the main function is in lowercase letters.

The official **CRAN** version can be installed with 

```{r, echo = TRUE, eval = FALSE}
install.packages("ANOFA")
library(ANOFA)
```

The development version `r packageVersion("ANOFA")` can be accessed through GitHub:

```{r, echo = TRUE, eval = FALSE}
devtools::install_github("dcousin3/ANOFA")
library(ANOFA)
```

The library is loaded with 

```{r, echo = TRUE, eval = FALSE, results = FALSE}
library(ANOFA)
```

# For more

As seen, the library `ANOFA` makes it easy to analyze frequency data.
Its general philosophy is that of ANOFAs.

The complete documentation is available on this 
[site](https://dcousin3.github.io/ANOFA/).

A general introduction to the `ANOFA` framework underlying this 
library can be found at *the Quantitative Methods for Psychology* @lc23b.

# References

\insertAllCited{}
