---
title: "Confidence intervals with frequencies"
bibliography: "../inst/REFERENCES.bib"
csl: "../inst/apa-6th.csl"
output: 
  rmarkdown::html_vignette
description: >
  This vignette describes how to plot confidence intervals with frequencies.
vignette: >
  %\VignetteIndexEntry{Confidence intervals with frequencies}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE, results = 'hide', warning = FALSE}
cat("this will be hidden; use for general initializations.\n")
library(ANOFA)
```

Probably the most useful tools for data analysis is a plot with
suitable error bars [@cgh21]. In this vignette, we show how
to make confidence intervals for frequencies

## Theory behind Confidence intervals for Frequencies

For frequencies, ANOFA does not
lend itself to confidence intervals. Hence, we decided to use
@cp34 pivot technique. This technique returns _stand-alone_
confidence intervals on a proportion, that is, an interval which can be 
used to compare an observed proportion to a theoretical value.
In order to compare an observed proportion to another proportion, it
is necessary to adjust the wide [see @cgh21]. Further, because we
want an interval for frequencies, we multiply the proportion by the 
total number of data $N$.

Herein, we use the @lt96 method which provides an analytical solution
based on the Fisher's $F$ distribution. It is based on $n$, the observed
frequency in a given cell, $N$, the total number of observations, and
$1-\alpha$, the confidence level (typically 95%). It is given by

$$\hat{\pi}_{\text{low}}=\left(
1+\frac{N-n+1}{n F_{1-\alpha/2}(2n,2(N-n+1)}
\right)^{-1}
< {\pi} <
\left( 
1+\frac{N-n}{(n+1) F_{\alpha/2}(2(n+1),2(N-x)}\right)^{-1}
=\hat{\pi}_{\text{high}}$$

This interval width is then multiply by $N$ with
$\{n_{\text{low}}, n_{\text{high}} \} = N \, \times\, \{ \hat{\pi}_{\text{low}}, \hat{\pi}_{\text{high}} \}$ 
so that we obtain an interval on frequencies rather than on probabilities.

Finally, the width of the intervals are adjusted for difference and 
correlations using:

$$n_{\text{low}}^* =  \sqrt{2} \sqrt{2}(n-n_{\text{low}})+n$$

$$n_{\text{high}}^* =  \sqrt{2} \sqrt{2}(n_{\text{high}}-n)+n$$ 

The lower and upper length are adjusted separately because this interval
may not be symmetrical (equal length) on both sides of the observed frequency $n$.

This is it.

## Complicated?

Well, not really:

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=3, fig.cap="**Figure 1**. The frequencies of the Light & Margolin, 1971, data as a function of aspiration for higher education and as a function of gender. Error bars show difference-adjusted 95% confidence intervals."}
library(ANOFA)
w <- anofa( obsfreq ~ vocation * gender, LightMargolin1971)
anofaPlot(w) 
```

You can select only some factors for plotting, with e.g., 

```{r, message=FALSE, warning=FALSE, fig.width=4, fig.height=3, fig.cap="**Figure 2**. The frequencies of the Light & Margolin, 1971, data as a function of gender. Error bars show difference-adjusted 95% confidence intervals."}
anofaPlot(w, obsfreq ~ gender) 
```

Because of the interaction _gender_ $\times$ _vocational aspiration_, 
the overall difference between boys and girls is
small. Actually, they differ only in their aspiration to go to the university
(recall that these are 1960s data...).

As is the case with any ``ggplot2`` figure, you can customize it at will. 
For example,

```{r, message=FALSE, warning=FALSE, fig.width=4, fig.height=3, fig.cap="**Figure 3**. Same as Figure 2 with some customization."}
library(ggplot2)
anofaPlot(w, obsfreq ~ gender) +
  theme_bw()
```


Here you go.


# References


